Avoidance in permutations

### 2014-11-07 project2/paper/paper.tex
Resolved TODO, and corrected typo

With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many four element sequences avoid 123? And more generally, how many $n$-element sequences avoid 123? If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$. In this paper we consider specific cases of a very general avoidance problem, describing the sequence $(s_n(\sigma))_n$ where $\sigma$ is any permutation of the elements of the set $\{1,2,\cdots, k\}\subset\mathbb{N}$ for some $k$.

@@

\begin{definition}
Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}\todo{I changed this definition... take a look and see if you agree with it}

@@

\begin{lemma}[Flipping Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Notice that we only need to proof one direction of this statement. The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation. We show that if ${\sigma}$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. Let $\pi=b_1\cdots b_k$. We proceed by induction. Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. Let $\sigma\in S_n$. If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.  Then $\sigma=a_1\cdots a_n$ where $a_i\in\{1,\cdots, n\}$. Without loss of generality, let $a_\ell=n$. Then the parent node of $\sigma$ is $\sigma_p=a_1\cdots a_{\ell-1} a_{\ell+1}\cdots a_n$. The induction hypothesis allows us to conclude that $\mathcal{F}(\sigma_p)$ avoids $\mathcal{F}(\pi)$. Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_p)$. Let $\mathcal{F}(\sigma)=a'_1\cdots a'_{\gamma-1} a_{\gamma}a'_{\gamma+1}\cdots a'_n$ and $\mathcal{F}(\sigma_p)=a'_1\cdots a'_{\gamma-1} a'_{\gamma+1}\cdots a'_n$ with $a'_\gamma=n$. Then $\mathcal{F}(\sigma_p)$ avoids $\pi$ while $\mathcal{F}(\sigma)$ does not. That means that there is a subsequence $\omega$ of $\mathcal{F}(\sigma)$ that has the same relative order as $\mathcal{F}(\pi)$. Let $\mathcal{F}(\pi)= b'_1\cdots b'_k$ with $b'_i=1$ and $b'_j=k$. So $\omega$ contains $a'_\gamma$, and $a'_\gamma$ occupies the same position as $b'_j$. Now if we remove 1 and $n$ from $\mathcal{F}(\sigma)$, then the result does not avoid $\mathcal{F}(\pi)$ with $b'_i=1$ and $b'_j=k$ omitted. Then if we apply a flip, by the induction hypothesis we see that the result of flipping $\mathcal{F}(\sigma)$ with $1$ and $n$ omitted does not avoid $\pi$ with $1$ and $k$ omitted. If we add in 1 and $n$ in the appropriate places to reconstruct $\mathcal{F}(\mathcal{F}(\sigma))=\sigma$ and if we reconstruct $\sigma$ similarly, then $\sigma$ does not avoid $\pi$, a contradiction. 
\end{proof}

>>>>

With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many $4$-element sequences avoid 123? And more generally, how many $n$-element sequences avoid 123? If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$. In this paper we consider specific cases of a very general avoidance problem, describing the sequence $(s_n(\sigma))_n$ where $\sigma$ is any permutation of the elements of the set $\{1,2,\cdots, k\}\subset\mathbb{N}$ for some $k$.

@@

\begin{definition}
Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}

@@

\begin{lemma}[Flipping Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Notice that we only need to prove one direction of this statement. The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation. We show that if ${\sigma}$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. Let $\pi=b_1\cdots b_k$. We proceed by induction.

Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. Let $\sigma\in S_n$. If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.  Then $\sigma=a_1\cdots a_n$ where $a_i\in\{1,\cdots, n\}$. Without loss of generality, let $a_\ell=n$. Then the parent node of $\sigma$ is $\sigma_p=a_1\cdots a_{\ell-1} a_{\ell+1}\cdots a_n$. The induction hypothesis allows us to conclude that $\mathcal{F}(\sigma_p)$ avoids $\mathcal{F}(\pi)$. Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_p)$. Let $\mathcal{F}(\sigma)=a'_1\cdots a'_{\gamma-1} a_{\gamma}a'_{\gamma+1}\cdots a'_n$ and $\mathcal{F}(\sigma_p)=a'_1\cdots a'_{\gamma-1} a'_{\gamma+1}\cdots a'_n$ with $a'_\gamma=n$. Then $\mathcal{F}(\sigma_p)$ avoids $\pi$ while $\mathcal{F}(\sigma)$ does not. That means that there is a subsequence $\omega$ of $\mathcal{F}(\sigma)$ that has the same relative order as $\mathcal{F}(\pi)$. Let $\mathcal{F}(\pi)= b'_1\cdots b'_k$ with $b'_i=1$ and $b'_j=k$. So $\omega$ contains $a'_\gamma$, and $a'_\gamma$ occupies the same position as $b'_j$. Now if we remove 1 and $n$ from $\mathcal{F}(\sigma)$, then the result does not avoid $\mathcal{F}(\pi)$ with $b'_i=1$ and $b'_j=k$ omitted. Then if we apply a flip, by the induction hypothesis we see that the result of flipping $\mathcal{F}(\sigma)$ with $1$ and $n$ omitted does not avoid $\pi$ with $1$ and $k$ omitted. If we add in 1 and $n$ in the appropriate places to reconstruct $\mathcal{F}(\mathcal{F}(\sigma))=\sigma$ and if we reconstruct $\sigma$ similarly, then $\sigma$ does not avoid $\pi$, a contradiction. 
\end{proof}
